---
title: "1_Introduction_to_R"
author: "Michael Tassia"
date: "`r Sys.Date()`"
output: html_document
---

## JHU Biology REU 2026 - Introduction on R

R is a common object-oriented programming language for performing data analysis in biology. While other programming languagues are also common in biology (e.g., python), R (and by extension, RStudio/Posit) is a common & flexible environment to simulate, analyze, and visualize biological data.

*Why program?* Programming presents a powerful way to interact and procedurally analyze/generate data in a reproducible manner (both in biology and other data-driven fields). In many subfields of biology (e.g., genetics and evolutionary biology), computer science has becoming inseparably intertwined with research. Becoming familiar/comfortable with one or two programming languages will provide far more flexibility when methods need to be catered to *your* data!

The goal of this 4-day workshop is to introduce the fundamentals of using R to analyze biological data. We'll start here with the basics of how to interact with the R programming environment and build off this foundation in the following lessons.

R and its libraries are frequently being updated, and new tools are being published every day to assist in data analysis. In this course, we'll only scratch the surface of computational biology. If you'd like further instruction, I recommend these resources as a starting point:

-   [Hands-On Programming with R](https://rstudio-education.github.io/hopr/)
-   [R for Data Science (2e)](https://r4ds.hadley.nz/)

Here, we'll begin by becoming familiar with the ***syntax*** of `R` (*how we interact with the programming environment*), common `R` ***objects*** (*what we interact with*), and how we manipulate objects in `R` using ***functions*** (*performing actions on objects*).

------------------------------------------------------------------------

## 1. The R syntax

To begin, we'll work through some examples of how to run code in `R`. Below, you'll find code blocks that you can run in your R environment. Some of these codeblocks are empty -- that's intentional! You can fill in the blanks to practice the concepts covered in this workshop.

Some important keyboard shortcuts:

-   `CTRL+ENTER` (Windows) or `CMD+RETURN` (Mac): *Run current line*
-   `CTRL+SHIFT+ENTER` (Windows) or `CMD+SHIFT+RETURN` (Mac): *Run current block*

Within the code blocks, you'll frequently see the `#` character. `#` tells R (as well as most other common programming languages) to ignore the rest of the line. This is a common way to add *comments* to your code -- describing what the code is doing, what the output means, or as guidance for your own workflow. For example, lets use the following intuitive code block to demonstrate how comments work:

```{r}
1 + 1 # Add 1 and 1 together
```

When running the code block above, you'll see that the output is `2` and the comment is ignored. Without the `#` character, the code block above will produce an error. We will use comments throughout this workshop to describe what the code is doing or provide instruction -- feel free to add your own comments too!

### 1.1 Operators

#### 1.1.1 Arithmetic, comparisons, creating vectors, and storing data

Here, we'll practice giving commands to R using common mathematical operators. A ***command*** is a line of code that tells `R` to perform some action on ***objects***. Let's apply some arithmetic ***operators*** to some ***integers*** below:

```{r}

# BASIC ARITHMETIC

10 + 5 # `+` sums
10 - 5 # `-` subtracts
10 / 5 # `/` divides
10 * 5 # `*` multiplies
10 ^ 5 # `^` exponentiate

# "COMPLEX" ARITHMETIC (follows PEMDAS)

10 + 5 - 10 * 5 # multiplication applied first
10 + (5 - 10) * 5 # parenthetical applied first

```

In the code block above, we used ***operators*** (i.e., `+`, `-`, `/`, `*`, `^`) to perform basic arithmetic tasks. In addition to arithmetic, operators can be used for many other tasks. We'll look at another three other types of operators below:

1.  *Comparisons* (e.g., `>`, `>=`, `==`, `!=`): These return a logical value (i.e., `TRUE` or `FALSE`) based on whether the comparison is true or false.
2.  *Creating vectors* using the `:` operator between two integers (e.g., `1:100` creates a vector of integers from 1 to 100, inclusive).
3.  *Storing data* using the `<-` operator (e.g., `example_integer <- 10` stores the integer `10` in the object `example_integer`).

```{r}
# 1. COMPARISONS

10 > 5 # Greater than
10 >= 5 # Greater than or equal to
10 == 5 # Equal to
10 != 5 # Not equal to

# 2. CREATING VECTORS

1:100 # Create a vector of all integers between (and including) 1 and 100
5*(1:100) # Arithmetic operators can be applied to integer series

# 3. STORING DATA

example_integer <- 10 # Save 10, an integer object, to the object `example_integer`
example_integer # Call the object `example_integer` to see its value

example_integer <- 10 * example_integer # Multiple the value of `example_integer` by 10 and store it in `example_integer`
example_integer # Call the object `example_integer` to see its new value

```

In the last example, we created a ***object*** stored in the string variable `example_integer` using the `<-` operator. An object stores data (e.g., an integer) or data structures (e.g., a matrix) using a string (e.g., `example_integer`). This task is fundamental to programming, as it allows you to save values and recall them later without needing to re-run the code that generated those values.

When you call an object by itself, the value of that object is often returned by printing (*note that this varies between classes of objects, but generally calling an object will either print its entirety to the screen or provide a brief summary*).

Importantly, it's recommended that you name objects intuitively, as they'll be important if you need to debug or recall specific values from your data. A few tips about naming objects:

-   Use `_` in place of spaces (e.g., `example_integer`)
-   Object names cannot begin with integers or operators (e.g., `1example_integer` and `!example_integer` do not work)
-   Capitol and lowercase letters are different (e.g., `example_integer` and `Example_integer` are two different objects)
-   Be consistent with your typesetting (e.g., `example_integer` & `results_output`, `EXAMPLE_INTEGER` & `RESULTS_OUTPUT`, `ExampleInteger` & `ResultsOutput`)

#### 1.1.2 Practice with operators {#practice-with-operators}

Now that we know the basics, let's practice using operators and objects with the following exercise. Below is a code block with tasks enumerated in comments. Try to fill in the blanks to complete the tasks (you can use the code above as a reference).

```{r}

# 1. Create an object that stores all integers from 1 to 100
  
# 2. For each value in the series, multiply each element by the sum of 10 and <2 raised to the power of 4>
  
# 3. For each value in the series, divide by the product of 10 and 5
  
# 4. How many values are greater than or equal to 50?

# 5. How many values in the object are less that 25?

```

### 1.2 Functions

***Functions*** are a type of object that store a series of commands. Function complexity ranges from simple tasks like computing a sum (i.e., `sum()`) to more complex statistical evaluations like performing linear regression (i.e., `lm()`). You can even write your own functions to perform specific tasks (we won't cover this here, but it's a powerful tool to have in your programming toolbox).

Functions have the following syntax: `function(<arguments>)` . ***Arguments*** can be thought of as "options" supplied to the function; each function may take zero or more arguments. Examples of arguments include the input data (e.g., `data = 0:100`) and ignoring missing values (e.g., `na.rm = TRUE`). Let's look at an example of a couple functions below:

```{r}

# Create a vector of integers from 1 to 100
dat <- 1:100

# Obtain the maximum value in the vector
max(dat)

# Obtain the minimum value in the vector
min(dat)

# Print a basic summary of the vector
summary(dat)

```

To learn more about a function, you can use the `help()` function to call the ***documentation*** about a function (e.g., `help(sum)`). A shorthand for this is to use the `?` operator for a function (e.g., `?sum`). The documentation for a function will often include a description of what the function does, the syntax for using the function, and examples of how to use the function.

Let's return to prompts in the [1.1.2 Practice with operators](#112-practice-with-operators) section above where we counted the number of values (stored in an object) that passed our test. Originally, we counted by eye; however, using `sum()`, we can quickly compute the number of values that pass our test. For example:

```{r}

# Using sum() with a logical vector will add the total number of `TRUE` values:
sum(1:50 > 25)

# Note that functions can also be nested, lets try an example below with `sqrt()` and `sum()`
sum(sqrt(1:50) > 5)

```

Applying ***functions*** to data stored in objects is at the core of how biological data is analyzed in R. At this point, we've covered the following concepts and we'll continuing using these tools in the upcoming lessons:

1.  ***Operators*** (e.g., `+`, `:`, `<-`)
2.  ***Objects*** (e.g., `example_integer`)
3.  ***Functions*** (e.g., `sum()`)

Next, we'll highlight some common types of objects and how to manipulate them.

------------------------------------------------------------------------

## 2. Common R objects - vectors and data frames

### 2.1 Vectors

We've already covered a few types of ***objects*** above (e.g., ***integers*** and ***functions***), so we'll begin with a object type we've already encountered: ***vectors***.

In R, ***vectors*** contain a one-dimensional data. Previously, we created a ***vector*** of ***integers*** using the `:` operator and the bounding values (e.g., `1:100`). However, ***vectors*** can be made of many different types of data using the `c()` function.

First, we'll create some vectors using the `c()` function:

```{r}

# 1. CREATING VECTORS
## 1.1 Create a character vector
example_vec <- c("Michael","Arjun","Margaret",
                 "Betty", "Josh", "Rajiv",
                 "Michael")
example_vec

## 1.2 Create an integer vector
example_vec <- c(2,4,6,8,10,12,14,16,18,20) # 2:20 by steps of 2
example_vec <- seq(2,20,2) # 2:20 by steps of 2; use ?seq to see this function's arguments
c(2,4,6,8,10,12,14,16,18,20) == seq(2,20,2) # Test whether each element of each vector is the same

```

Using the `example_vec` created above, let's practice some common ways to work with ***vectors***. Note that the examples below also apply to other types of vectors (e.g., character vectors, logical vectors).

```{r}

# 2. COMMON FUNCTIONS
length(example_vec) # Return number of elements in vector
head(example_vec, n=5) # Show first 5 elements
tail(example_vec, n=5) # Show last 5 elements

# 3. ACCESSING DATA
example_vec[1] # Access first element
example_vec[6] # Access sixth element
example_vec[length(example_vec)] # Access last element
example_vec[example_vec >= 12] # Access values > 12

# 4. EDITING DATA
example_vec <- seq(0,100,5) # Create an integer vector containing 0-100 by steps of 5

example_vec[1] <- 10e6 # Replace the first value with 10e6
example_vec <- c(example_vec,0) # Add 0 to the end of the vector
example_vec[2] <- NA # Replace the second value with NA
example_vec[example_vec == 50] <- 500 # Replace any 50 with 500
example_vec

# COMPLEX EXAMPLE - What are we filtering for here?
example_vec[example_vec > 15 & example_vec <= 40 & !is.na(example_vec)]

```

***Vectors*** are commonly used to in R, so being familiar with how they can be filtered, accessed, and manipulated is very powerful. Though we've only scratched the surface here, we'll be using ***vectors*** frequently throughout this workshop and you'll likely encounter ***vectors*** if you perform data analysis in R.

#### 2.1.1 Practice with vectors

Let's try an exercise [*without*]{.underline} focusing on a ***vector*** of integers. Unlike integers (e.g., `1, 2, 3, 4, ...`), characters in R are specified by quotations around the string, such as `"salamander"`. Thus, `1` and `"1"` are entirely different objects.

In this example, we'll specify the ***functions*** and ***arguments*** that can be used (remember, you can use `help(<function name>)` or `?<function name>` for documentation):

```{r}

letters # Pre-existing vector of english letters

# 1. Use sample() to create a vector object with the following traits:
#   - Contains 100 random samples from the `letters` object
#   - Sample letters with replacement (each time you pick a letter, that letter needs to be able to be picked again)
#   - NOTE: Store your object in a variable called something other than `letters`

# 2. Using your vector object, perform the following tasks (in order):
#   a. Count each letter that appears in your first name
#   b. Replace each occurrence of 'z' with 'x'
#   c. Add each letter in your name to the end of your character vector
#   d. Use `sort()` on your object, then print the 50th - 60th elements. 

```

Here, we're working with random letters of the alphabet. However, these types of procedures can also be used to filter and access, for example, specific genome loci stored in a *variant call file (VCF)*.

### 2.2 Data frames & R packages

Perhaps the most common way to collect, store, and share data is in the format of a two-dimensional "table" organized in rows (observations) and columns (variables/features). If you're familiar with an `Excel Spreadsheet`, then you're familiar with R's ***data frame objects***. For example, let's look at a ***data frame*** stored in the `datasets` ***package.***

```{r}

datasets::iris

```

#### 2.2.1 Loading packages

In this section, we'll focus on how to generate, access, and manipulate ***data frames*** - the most common format to store data in `R`. We'll also introduce and use ***packages*** (such as `datasets`) which are a collection of useful tools, instructions, and data that are purpose-built to assist users in specific analyses (e.g., phylogeny or generalized linear models). In this lesson, we'll be using the `datasets` ***package*** to access some ***data frames*** to practice what we've learned. In later lessons, we'll use a powerful suite of packages called `tidyverse` which helps organize and visualize data (e.g., plotting) of many different formats.

In the example above, we called the `iris` ***data frame*** from the `datasets` ***package*** using the `::` operator. The `::` operator allows you to access an ***object*** from a ***package*** (in the example above, we accessed a ***data frame***, though the same principle applies to using ***functions*** from a ***package***). The syntax for the `::` operator is: `<package>::<object>`.\*

As you may imagine, needing to call the ***package*** every time we use it would quickly bloat our code. Fortunately, a ***package*** can be loaded into your coding environment using the `library()` ***function***, allowing you to succinctly call ***objects*** without specifying their source (just as you had with `sum()` or `sample()`). Here are a few notes on how to load ***packages***, unload ***packages***, and receive help:

```{r, results=FALSE, message=FALSE, warning=FALSE}

library() # list all available packages for loading
library("datasets") # load the `datasets` package into the environment

library(help = "datasets") # load the documentation for the package
detach("package:datasets") # unload the package

```

#### 2.2.2 Working with data frames

Now that we know about ***packages***, let's return to working with ***data frames***. First, let's focus on how we interact with ***data frames***. Some of the syntax below will be familiar to how we interacted with ***vectors***.

```{r}

# Load library and iris data frame
library("datasets") # Load the `datasets` package
df<-iris # Create a copy of the `iris` dataframe

# Print the first six lines of the dataframe
head(df)
```

In our examples of accessing data in ***vectors***, we used the `[]` operator to access specific values stored within the ***vector*** (e.g., `example_vec[5]` to access the 5th element). This syntax worked because a ***vector*** is one-dimensional.

Because a ***data frame*** is two-dimensional, R interprets the values inside the brackets differently. Here's how integers in brackets work for accessing values in a ***data frames:***

-   ***Integers*** without a comma will return whole columns: `df[1:3]`
-   ***Integers*** followed by a comma will return rows: `df[1:3,]`
-   ***Integers*** preceded by a column will extract columns as a vector: `df[,1:3]`

It follows then, to access a subset of a ***data frame*** you can use the following syntax: `df[<rows>,<columns>]`, such as `df[1:3,2:4]` to access rows `1:3` and columns `2:4`. You can also use ***vectors*** to reorganize your ***data frames*** (e.g., `df[1:10,c(5,3)]`). As with most ***objects*** in R, changes to an ***object*** are not saved unless you write those changes to an ***object.***

Another important element of ***data frames*** is that the columns are often associated with a character string (i.e., a *column* *name*). In the following examples, we'll see how we can access and filter a ***data frame***:

```{r}

# Note that we don't have to reload the `datasets` library as it's still active in our current `R` environment. 

# Create a copy of the `iris` data frame 
df <- iris 

# 1. VIEWING DATA
df # Print `df`

head(df) # Print first part of `df`
tail(df) # Print last part of `df`

colnames(df) # Print column names of `df`
rownames(df) # Print row names of `df`

nrow(df) # Print number of rows in `df`
ncol(df) # Print number of columns in `df`
```

```{r}

# 2. EXTRACTING DATA
df[,1] # Extract the first column
df$Sepal.Length # Extract the column named "Sepal.Length"
setequal(df[,1], df$Sepal.Length) #Test whether two vectors are equal

df[,c("Species","Sepal.Length")] # Extract multiple columns by name
df[df$Species == "virginica",] # Extract rows by testing the value of a column

# 3. EDITING DATA
df <- iris # Reset the `df` object to `iris`
df[nrow(df)+1,] <- c(100,100,100,100,NA) # Append a row (observation)
df$collection_month <- rep("June",times=nrow(df)) # Append a column (variable)
df
```

With these methods in hand, what does the following code do?

```{r}
# Create a new copy of the `iris` data set
df <- iris

df$big <- FALSE
df[df$Species == "setosa" & df$Petal.Length > 1.6,]$big <- TRUE
df[df$Species == "virginica" & df$Petal.Length > 6.0,]$big <- TRUE
df[df$Species == "versicolor" & df$Petal.Length > 4.5,]$big <- TRUE
summary(df[df$big == TRUE,]$Species)
```

As you can see, increasing the dimensions of our data from one (***vector***) to two (***data frame***) increased the complexity of working with such data. The `iris` data set as an example of a very common format used to record data - particularly if your work involves fieldwork (no pun intended). In fact, having the foresight to record data in a format similar to how you store data can save a lot of time in the future!

Before we move onto the final example exercise, let's walk through how to create a ***data frame*** from scratch using simulated data. Below, we'll use a few common ***functions*** to simulate (or randomly sample) data. We'll also use some formatting tricks that will help with making code a bit easier to read. If you're not sure what a ***function*** is doing below, try running each ***function*** by itself in the console below (and looking at that ***function's*** `help()`).

```{r}

df<-data.frame(trial = rep(x = 1:50, # `rep()` replicates values of a vector
                           each = 2),
               species = rep(x = c("a","b"),
                             times = 50),
               height = rnorm(n=100, # `rnorm()` simulates values from a normal distribution
                              mean=20,
                              sd=5),
               color = sample(x = c("blue","red","green"),
                              size = 100,
                              replace = T))

colnames(df)
nrow(df)
summary(df)
```

While empirical data analysis (i.e., analyzing observational data) may be the more immediately relevant use-case for R, simulation of data (e.g., output from `rnorm()`) is *extremely* important for hypotheses testing and generating null expectations. We return to this concept in a later lesson.

#### 2.2.3 Practice with data frames

Let's close out by working through a ***data frame*** exercise that uses everything we've learned.

```{r}

# Load the `airquality` dataset into a data frame object
df <- airquality

# Count the rows missing (NA) an Ozone measurement
# Hint: Remember the function is.na() 


# Remove all observations from the months of May and June (you may need to do this in multiple steps)
# Hint: You can use `!` to negate a comparison 

# How many days was Wind greater than 10?

# Create a column called "observer" that contains your name

# Look at the 20th line of your data frame 

```

Congratulations! We've covered the very basics of how to code in `R`. We've only scratched the surface, but you've now encountered many of the core elements of what makes R such a powerful tool for data science!

Above, we covered several examples of the ***operators*** and ***functions*** used to interact with ***objects***. In the next lesson, we'll introduce the `tidyverse` ***package*** (for data tidying and plotting) and work with some basic summary statistics used in data analysis.

------------------------------------------------------------------------

## 3. Recommended reading

-   [Chapters 2 in ***Hands on Programming with R***](https://rstudio-education.github.io/hopr/basics.html)
-   [*Chapter 2 in **R for Data Science***](https://r4ds.hadley.nz/workflow-basics)
